Refinements for open automata

### 2022-07-11 2021-Refinement/paper/refinementarticle.tex
Suggestions de modifications

@@

Establishing equivalence and refinement relations between programs is an important mean for 
verifying their correctness. By establishing that the
behaviours of a modified program simulate those of the source one,  and bisimulation relations  formalise the desired  relation between a specification and an implementation, 
 two equivalent implementations, or a program and its optimised implementation.
This article discusses a notion of refinement between \emph{open automata}, which are symbolic
behavioural models for communicating systems. 
Open automata may have \emph{holes} modelling elements of their
context, and can be composed by instantiation of the holes. This allows for a compositional approach for
verification of their behaviour.

@@

The notion of refinement  captures the relation between  a specification and an implementation of the same component. Refinement entails that one system can be considered as a more precise version of the specification, featuring all the specified behaviours with more concrete details. 
 It is usually defined as trace inclusion or simulation  \cite{Milner:1980,Kouchnarenko:2007}; this ensures that all behaviours of the implementation must be also behaviours of the specification. \TODO{heu c'est pas le contraire, l'implem (le "refined"), peut avoir plus de comportements que la spec, non ?}
 This definition, which is based on systems whose behaviour is fully defined, is not well-suited for open systems,  as it requires to reason also about unspecified behaviours.
In this article we mostly rely on a simulation point of view, where all the specified behaviours must be followed by the refined system but additional behaviours may exist. \TODO{the following sentence is a copy of the previous paragraph !} However we also ensure that a whole scenario, made of several steps, of the specification can also be simulated by the refined system, which is a bit richer than the traditional simulation relation.

Open automata were defined as a way to provide a semantics for open parameterized hierarchical LTSs used in verification tools.
An open automaton \cite{henrio:01299562} is a classical automaton with variables and holes. Variables make automata symbolic and allow them to encode infinite state systems. Holes enable the composition of automata: an automaton with a hole is an operator that takes another automaton as parameter and reacts to the actions it emits; the composed automaton is a more precise automaton where the behaviour of one ``parameter'' \TODO{est-ce qu'on n'ecrirait pas ici "process parameter" pour les distinguer des variables standard ?} of the main automaton has been provided.

@@

%\TODO{useful or not?}
\begin{definition}[Out-transition, Transition variables]
Let \(V\) be the variable names of the considered automaton, \(T\) its transitions and \(r\) one of its states.
\(\fOT[T]{r} \in T\) are called the out-transitions of the state \(r\).
%\(\fIT[T]{r} \in T\) are called the in-transitions of the state \(r\).
When the transition set is clear from the context, it will be omitted.
The local variables of a transition \(\fvars{t}\) are all variables appearing in transition \(t\) except the variables of the automaton.
%\begin{align*}
% %&
%%	\fIT[T]{r} & = \mset{\OTg \in T}{s' = r} \\
%\end{align*}
%\vspace{-1cm}
\begin{mathpar}
%	\fguard{\OTg} \!=\! g \qquad
	\fOT[T]{r}  = \mset{\OTg \in T}{s = r}

@@

\begin{definition}[Composition of open automata] \label{Def:CompOA}
The composition of \(A_c = \OAg[c]\) in the hole \(k \in J_c\) of  \(A_p = \OAg[p]\) is the OA defined as follows:
\[A_c\subst{A_p}{k} ::=  \OA{S_c \times S_p}{\mpar{s_{0c}, s_{0p}}}{V_c \uplus V_p}{\sigma_{0c} \uplus \sigma_{0p}}{J_p \uplus J_c \setminus \mbrc{k}}{T} \] \text{with }

@@

\paragraph{Relations between open automata}
A relation (bisimulation, refinement, etc.) between OA requires to compare their states. To do so we will suppose that the variables of the two OA are disjoint (a renaming of variables may have to be applied before comparing OA states).
\begin{definition}[Relation on OA states] Suppose $V_1$ and $V_2$ are disjoint.
A relation on states of \(\OAg[1]\) and \(\OAg[2]\) is a function \(R: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\).
\end{definition}
\TODO{C'est etrange, la formulation "the variables they refer to", comme il n'y a plus de variables d'etat...}
The idea is that two states are related if the variables they refer to verify a certain formula. Additionally, we may check that initial states of the automata are related by checking that: \(\sigma_{01} \uplus \sigma_{022} \vdash R\mpar{s_{01}, s_{02}}\).

@@

Similarly to FH-bisimulation~\cite{fhbisim} we are interested  in finding relations between states of two open automata that contain variables and holes. However here we want to build a refinement relation that  also guarantees that no deadlock is introduced when refining the automaton (at least not among the scenarios originally planned in the specification).
The refinement relation should be conditioned by the internal states of the automata. More precisely, a refinement relation characterizes when two states are related, and this  characterisation is expressed as a predicate on the variables of the two automata.
Thus, a refinement relation has the following signature: \( S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\).

@@

Let us first explain the reasons why simulation can introduce deadlocks and how we want too limit them. 
The automaton that simulates a specification features any behaviour that contains the behaviour of the specification, or more precisely, for each state that simulates a state of the specification, the simulating automaton must at least feature the behaviour of the specification.
 In this definition, nothing prevents the implementation from adding non-determinism and non-deterministically reaching a state that is unrelated to the specification, this state could be a deadlock even though we only followed transitions that exist in the specification.
To avoid the appearance of such deadlocks, we will state that, when both the implementation and the specification do the same transition, they stay related.
This ensures a form of continuity in the comparison in the sense that the refined system is constrained to follow the behaviour of the simulated system along a given scenario. 
In terms of traces this amounts to some form of deadlock reduction: for each trace followed by the simulated system if the refined system followed the beginning of the trace, it must be able to keep following the trace. 
Of course, the classical trace view is a simplified view compared to the guards, states and actions from  the holes that exist in OA, but it gives the right intuition.

@@

We thus define the following property that can be understood as follows. Take a specification automaton and its implementation. Consider two related states for the refinement relation. Suppose an identical transition exist in both automata (with same guards and same label), then either the reached states $s_1'$ and $s_2'$ are related by refinement, or there exists another transition of the specification that is also identical and leads to a state $s_1''$ that is related to $s_2'$ in the refinement relation.
Finally, we should take into account that 1) instead of equality, we should check if there is a state such that both transitions are feasible (the conjunction of guards is satisfiable), and that  2)instead of finding another transition of the specification we might find a family of transitions that cover all the cases of the transition of the implementation. This leads us to the following definition.
\begin{definition}[Preservation along identical transitions]\label{def:preservation} \\
Let \(\OAg[1]\) and \(\OAg[2]\) be two OAs.
A relation \(R: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\) between  \(\OAg[1]\) and \(\OAg[2]\)
  is preserved along identical transitions if  it satisfies the following:
\begin{multline*}
\forall \mpar{s_1, s_2} \in S_1 \times S_2, \bigsymb{\forall} \mpar{t_1, t_2} = \mpar{\OTx{1}{}{1}{1}, \OTx{2}{}{2}{2}} \in \fOT{s_1} \times \fOT{s_2} \\ 
\bigsymb{\bigexists} 
\mpar{t_{1x} = \OTx{1}{x}{1x}{1x} \in \fOT{s_1}}^{x \in X},
 V_1 \uplus V_2 \uplus \biguplus_{x\in X}\fvars{t_{1x}} \uplus \fvars{t_2} \uplus \fvars{t_1} \vdash \hspace{6cm}\\  R\mpar{s_1, s_2} \wedge 	g_1 \wedge g_2 \wedge \alpha_1 = \alpha_2 
%\wedge \bigwedge_\subbox{j \in J'_1 \cap J'_2} \beta_{1j} = \beta_{2j} 
\implies \\
%\exists
%\OTx{1}{a}{3}{3}\in \fOT{s_1}. \\
\bigsymb{\bigwedge_{x\in X}}
\mpar{
\alpha_1=\alpha_{1 x}  \land J'_{1x}= J'_2 \land \bigwedge_\subbox{j \in J'_2 } \beta_{2j} = \beta_{1xj}\land g_{1 x}\land
 R\mpar{s'_{1 x}, s'_2}\psubst{\psi_{1 x} \uplus \psi_2}}\\
\end{multline*}
\end{definition}

@@

This section  extends the preceding relation to automata where the set of holes is not the same. A simple use-case for this is filling a hole with a completely defined automaton. In this case, we want to ensure that the automaton with a filled hole is a refinement of the other automata: actions of the identical holes will be taken into account the same way, and filling the hole partially reduces the behaviour of the automaton.

The major challenge in the design of this relation is to maintain a form of transitivity while being able to take into account the actions of some of the holes. A naive definition of refinement would ensure that the holes that are identical in the two open automata are taken into account in the simulation. Unfortunately considering all the common holes does not ensure transitivity of the simulation for the following reason. If $A_1$ simulates $A_2$ and $A_2$ simulates $A_3$, and one hole $j$ appears in $A_3$ and in $A_1$ but not in $A_2$ then we have no guarantee on the way $A_1$ and $A_3$ take the actions of this his hole into account, thus  a refinement between and $A_1$ and $A_3$ would require conditions involving actions of the hole $j$ which cannot be ensured. The way we solve this issue is to remember in the simulation relation which holes have been compared. This way the relation is parameterized by the set of holes that belong to the two automata and are taken into account.
This way, in the example above, we would have no guarantee on actions the hole $j$ by transitivity but can state a refinement relation with guarantees on the actions of the other holes.

@@

The relation  \(\wrel{}{}{H}\) is reflexive,  \(\wrel{A}{A}{H}\),  by taking $R\mpar{s_1, s_2} \mapsto \displaystyle\bigwedge_\subbox{v \in vars(s_1)} {v=v}$ to be $s_1= s_2$  and checking the above conditions (Definition \ref{Def:OA-Refinement}), we can see that the relation is indeed reflexive.
%
%\begin{theorem}[Transitivity]
%If $\wrel{A_1}{A_2}{H}$ and $\wrel{A_2}{A_3}{H'}$, then $\wrel{A_1}{A_3}{H\cap H'}$.
%\end{theorem}
Appendix~\ref{sec:proof-transitivity} presents the proof of transitivity. It is done classically by identifying the relation between $A_1$ and $A_3$ that is a refinement. What is less classical is the definition of this relation because it is a boolean formula. For each couple of states  $s_1$ and $s_3$ of $A_1$ and $A_3$ we build a a formula that defines the refinement relation. To do this, we take the disjunction of formulas relating $s_1$ and $s_3$, and passing by all states $s_2$ of $A_2$. More precisely, we define a relation of the following form:
  \[R_{13}(s_1,s_3)=\bigvee_{s_2\in S_2}\mpar{R_{12}(s_1,s_2) \land R_{23}(s_2,s_3) } \]
We then prove that this relation defines is a refinement, according to Definition~\ref{Def:OA-Refinement}.

@@

\begin{center}
\includegraphics[scale=0.7]{Figures/Theorem1}
\end{center}

Open automata composition is a refinement where
the tracked holes are the holes of the base automaton minus the composed hole.
\begin{theorem}[Composition is a refinement]\label{thm:Composition}
Let $A_1$ and  $A_2$  be three two automata with $J_1$ the set of holes of $A_1$.
Suppose that \(k \in J_1\) and that \(A_1\subst{A_2}{k}\) is non-blocking. We have:
\[ \wrel{A_1\subst{A_2}{k}}{A_1}{J_1\setminus \{k\}}\]
\end{theorem}

>>>>

% TODO: would you like some consistency work in the notations of transitions (esp. True formula and empty map)

@@

Establishing equivalence and refinement relations between programs is an important mean for 
verifying their correctness. By establishing that the
behaviours of a modified program simulate those of the source one, bisimulation relations formalise the desired relation between a specification and an implementation, 
 two equivalent implementations, or a program and its optimised implementation.
This article discusses a notion of refinement between \emph{open automata}, which are symbolic
behavioural models for communicating systems. 
Open automata may have \emph{holes} modelling elements of their
context, and can be composed by instantiation of the holes. This allows for a compositional approach for
verification of their behaviour.

@@

The notion of refinement  captures the relation between  a specification and an implementation of the same component. Refinement entails that one system can be considered as a more precise version of the specification, featuring all the specified behaviours with more concrete details. 
 It is usually defined as trace inclusion or simulation  \cite{Milner:1980,Kouchnarenko:2007}; this ensures that all behaviours of the implementation must be also behaviours of the specification. \TODO{heu c'est pas le contraire, l'implem (le "refined"), peut avoir plus de comportements que la spec, non ? +1}
 This definition, which is based on systems whose behaviour is fully defined, is not well-suited for open systems,  as it requires to reason also about unspecified behaviours.

Open automata were defined as a way to provide a semantics for open parameterized hierarchical LTSs used in verification tools.
% TODO? This could be an opportunity to introduce pNets as they are reffered to later but not introduced
An open automaton \cite{henrio:01299562} is a classical automaton with variables and holes. Variables make automata symbolic and allow them to encode infinite state systems. Holes enable the composition of automata: an automaton with a hole is an operator that takes another automaton as parameter and reacts to the actions it emits; the composed automaton is a more precise automaton where the behaviour of one ``parameter'' \TODO{est-ce qu'on n'ecrirait pas ici "process parameter" pour les distinguer des variables standard ?} of the main automaton has been provided.

@@

%\TODO{useful or not?}
\begin{definition}[Out-transition, Transition variables]
Let \(V\) be the variable names of the considered automaton, \(T\) its transitions and \(r\) one of its states.
\(\fOT[T]{r} \subset T\) are called the out-transitions of the state \(r\).
%\(\fIT[T]{r} \subset T\) are called the in-transitions of the state \(r\).
When the transition set is clear from the context, it will be omitted.
The local variables of a transition \(\fvars{t}\) are all variables appearing in transition \(t\) except the variables of the automaton.
%\begin{align*}
% %&
%%	\fIT[T]{r} & = \mset{\OTg \in T}{s' = r} \\
%\end{align*}
%\vspace{-1cm}
\begin{mathpar}
%	\fguard{\OTg} \!=\! g \qquad
	\fOT[T]{r}  = \mset{\OTg \in T}{s = r}

@@

\begin{definition}[Composition of open automata] \label{Def:CompOA}
% TODO: this definition is a bit strange, why is Ac in the hole of Ap but the hole is in Ac and the notation is reversed?
The composition of \(A_c = \OAg[c]\) in the hole \(k \in J_c\) of  \(A_p = \OAg[p]\) is the OA defined as follows:
\[A_c\subst{A_p}{k} ::=  \OA{S_c \times S_p}{\mpar{s_{0c}, s_{0p}}}{V_c \uplus V_p}{\sigma_{0c} \uplus \sigma_{0p}}{J_p \uplus J_c \setminus \mbrc{k}}{T} \] \text{with }

@@

\paragraph{Relations between open automata}
A relation (bisimulation, refinement, etc.) between OA requires to compare their states. To do so we will suppose that the variables of the two OA are disjoint (a renaming of variables may have to be applied before comparing OA states).
\begin{definition}[Relation on OA states] Suppose $V_1$ and $V_2$ are disjoint.
A relation on states of \(\OAg[1]\) and \(\OAg[2]\) is a function \(R: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\).
\end{definition}
\TODO{C'est etrange, la formulation "the variables they refer to", comme il n'y a plus de variables d'etat... +1}
The idea is that two states are related if the variables they refer to verify a certain formula. Additionally, we may check that initial states of the automata are related by checking that: \(\sigma_{01} \uplus \sigma_{02} \vdash R\mpar{s_{01}, s_{02}}\).

@@

Similarly to FH-bisimulation~\cite{fhbisim} we are interested  in finding relations between states of two open automata that contain variables and holes. However here we want to build a refinement relation that  also guarantees that no deadlock is introduced when refining the automaton (at least not among the scenarios originally planned in the specification).
The refinement relation should be conditioned by the internal states of the automata. More precisely, a refinement relation characterizes when two states are related, and this  characterisation is expressed as a predicate on the variables of the two automata.
Thus, a refinement relation has the following signature: \( S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\). % TODO: duplicata avec Def Relation on OA states

@@

% TODO: simulation doesn't introduce deadlocks, it doesn't forbid it, or tolerates it but it doesn't introduces deadlocks by itself.
Let us first explain the reasons why simulation can introduce deadlocks and how we want too limit them. 
The automaton that simulates a specification features any behaviour that contains the behaviour of the specification, or more precisely, for each state that simulates a state of the specification, the simulating automaton must at least feature the behaviour of the specification.
 In this definition, nothing prevents the implementation from adding non-determinism and non-deterministically reaching a state that is unrelated to the specification, this state could be a deadlock even though we only followed transitions that exist in the specification.
To avoid the appearance of such deadlocks, we will state that, when both the implementation and the specification do the same transition, they stay related.
This ensures a form of continuity in the comparison in the sense that the refined system is constrained to follow the behaviour of the simulated system along a given scenario. 
In terms of traces this amounts to some form of deadlock reduction: for each trace followed by the simulated system if the refined system followed the beginning of the trace, it must be able to keep following the trace. 
Of course, the classical trace view is a simplified view compared to the guards, states and actions from  the holes that exist in OA, but it gives the right intuition.

@@

We thus define the following property that can be understood as follows. Take a specification automaton and its implementation. Consider two related states for the refinement relation. Suppose an identical transition exist in both automata (with same guards and same label), then either the reached states $s_1'$ and $s_2'$ are related by refinement, or there exists another transition of the specification that is also identical and leads to a state $s_1''$ that is related to $s_2'$ in the refinement relation.
Finally, we should take into account that 1) instead of equality, we should check if there is a state such that both transitions are feasible (the conjunction of guards is satisfiable), and that  2)instead of finding another transition of the specification we might find a family of transitions that cover all the cases of the transition of the implementation. This leads us to the following definition.
\begin{definition}[Preservation along identical transitions]\label{def:preservation} \\
Let \(\OAg[1]\) and \(\OAg[2]\) be two OAs.
A relation \(R: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\) between  \(\OAg[1]\) and \(\OAg[2]\)
  is preserved along identical transitions if  it satisfies the following:
% TODO: isn't it a2 = a1x instead of a1x = a1
\begin{multline*}
\forall \mpar{s_1, s_2} \in S_1 \times S_2, \bigsymb{\forall} \mpar{t_1, t_2} = \mpar{\OTx{1}{}{1}{1}, \OTx{2}{}{2}{2}} \in \fOT{s_1} \times \fOT{s_2} \\ 
\bigsymb{\bigexists} 
\mpar{t_{1x} = \OTx{1}{x}{1x}{1x} \in \fOT{s_1}}^{x \in X},
 V_1 \uplus V_2 \uplus \biguplus_{x\in X}\fvars{t_{1x}} \uplus \fvars{t_2} \uplus \fvars{t_1} \vdash \hspace{6cm}\\  R\mpar{s_1, s_2} \wedge 	g_1 \wedge g_2 \wedge \alpha_1 = \alpha_2 
%\wedge \bigwedge_\subbox{j \in J'_1 \cap J'_2} \beta_{1j} = \beta_{2j} 
\implies \\
%\exists
%\OTx{1}{a}{3}{3}\in \fOT{s_1}. \\
\bigsymb{\bigwedge_{x\in X}}
\mpar{
\alpha_1=\alpha_{1 x}  \land J'_{1x}= J'_2 \land \bigwedge_\subbox{j \in J'_2 } \beta_{2j} = \beta_{1xj}\land g_{1 x}\land
 R\mpar{s'_{1 x}, s'_2}\psubst{\psi_{1 x} \uplus \psi_2}}\\
\end{multline*}
\end{definition}

@@

This section  extends the preceding relation to automata where the set of holes is not the same. A simple use-case for this is filling a hole with a completely defined automaton. In this case, we want to ensure that the automaton with a filled hole is a refinement of the other automata: actions of the identical holes will be taken into account the same way, and filling the hole partially reduces the behaviour of the automaton.
% TODO: what does this last part (after comma) means?

The major challenge in the design of this relation is to maintain a form of transitivity while being able to take into account the actions of some of the holes. A naive definition of refinement would ensure that the holes that are identical in the two open automata are taken into account in the simulation. Unfortunately considering all the common holes does not ensure transitivity of the simulation for the following reason. If $A_1$ simulates $A_2$ and $A_2$ simulates $A_3$, and one hole $j$ appears in $A_3$ and in $A_1$ but not in $A_2$ then we have no guarantee on the way $A_1$ and $A_3$ take the actions of this his hole into account, thus  a refinement between and $A_1$ and $A_3$ would require conditions involving actions of the hole $j$ which cannot be ensured. The way we solve this issue is to remember in the simulation relation which holes have been compared. This way the relation is parameterized by the set of holes that belong to the two automata and are taken into account.
This way, in the example above, we would have no guarantee on actions the hole $j$ by transitivity but can state a refinement relation with guarantees on the actions of the other holes.
% TODO: Too many "way"

@@

% TODO: sorry, what is R? I don't understand this definition, why v = v on vars(s1), what about s2
The relation  \(\wrel{}{}{H}\) is reflexive,  \(\wrel{A}{A}{H}\),  by taking $R\mpar{s_1, s_2} \mapsto \displaystyle\bigwedge_\subbox{v \in vars(s_1)} {v=v}$ to be $s_1= s_2$  and checking the above conditions (Definition \ref{Def:OA-Refinement}), we can see that the relation is indeed reflexive.
%
%\begin{theorem}[Transitivity]
%If $\wrel{A_1}{A_2}{H}$ and $\wrel{A_2}{A_3}{H'}$, then $\wrel{A_1}{A_3}{H\cap H'}$.
%\end{theorem}
Appendix~\ref{sec:proof-transitivity} presents the proof of transitivity. It is done classically by identifying the relation between $A_1$ and $A_3$ that is a refinement. What is less classical is the definition of this relation because it is a boolean formula. For each couple of states  $s_1$ and $s_3$ of $A_1$ and $A_3$ we build a a formula that defines the refinement relation. To do this, we take the disjunction of formulas relating $s_1$ and $s_3$, and passing by all states $s_2$ of $A_2$. More precisely, we define a relation of the following form:
% TODO: this relation defines is
  \[R_{13}(s_1,s_3)=\bigvee_{s_2\in S_2}\mpar{R_{12}(s_1,s_2) \land R_{23}(s_2,s_3) } \]
We then prove that this relation defines is a refinement, according to Definition~\ref{Def:OA-Refinement}.

@@

% TODO: why is there several ovoid stacked only on one side?
\begin{center}
\includegraphics[scale=0.7]{Figures/Theorem1}
\end{center}

% TODO: non-blocking is not defines afaik
Open automata composition is a refinement where
the tracked holes are the holes of the base automaton minus the composed hole.
\begin{theorem}[Composition is a refinement]\label{thm:Composition}
Let $A_1$ and  $A_2$  be three two automata with $J_1$ the set of holes of $A_1$.
Suppose that \(k \in J_1\) and that \(A_1\subst{A_2}{k}\) is non-blocking. We have:
\[ \wrel{A_1\subst{A_2}{k}}{A_1}{J_1\setminus \{k\}}\]
\end{theorem}
